---
title: "Saccharomycotina Yeast Phylogenetics Pipeline"
subtitle: "Mannose-associated conserved genes in 267 yeast species"
author: "Benjamin Narh-Madey"
date: "March 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: show
    code-tools: true
    theme: flatly
    highlight-style: github
    embed-resources: true
    fig-width: 10
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# Core packages — install if missing
pkgs <- c("ape", "caper", "picante", "phytools", "ggplot2", "dplyr",
          "tidyr", "knitr", "kableExtra")
new_pkgs <- pkgs[!pkgs %in% installed.packages()[,"Package"]]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cloud.r-project.org")

suppressPackageStartupMessages({
  library(ape)
  library(caper)
  library(picante)
  library(phytools)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(knitr)
  library(kableExtra)
})

# Set working directory to project root — adjust this path as needed
# PROJECT_ROOT <- "/home/glbrc.org/narhmadey/2.Y1000_267_phylogenetics/2_Plant_Path_563"
# setwd(PROJECT_ROOT)

knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%"
)
```

## Project Overview

This pipeline analyzes **267 focal yeast species** from the Y1000+ *Saccharomycotina* dataset to identify gene families (orthogroups) conserved in mannose-utilizing species compared with non-utilizers. Machine learning models first identified orthogroups (OGs) associated with the mannose growth phenotype. The pipeline then aligns, trims, and infers gene trees for each orthogroup, followed by phylogenetic signal analysis.

**Two-stage workflow:**

| Stage | Where | Steps |
|-------|-------|-------|
| Preprocessing | Local / submit node | Subset FASTAs → MAFFT → TrimAl → IQ-TREE (test) |
| Analysis | GLBRC HT Condor cluster | ModelFinder + IQ-TREE trees → RAxML-NG → ASTRAL → BEAST2 → D-statistic (R) |

**How the IQ-TREE trees flow through the pipeline:**

```
Phase 1 (IQ-TREE -m MFP -B 1000)
    │
    ├─── jobs_with_models.txt  ──────────────────────► Phase 2a (RAxML-NG --model)
    │                                                       │
    └─── iqtree_logs/OG*.treefile  ─────────────────► Phase 2b (BEAST2 starting tree)
                                                            │
Phase 2a (RAxML-NG bestTree / support)                     │
    │                                                       │
    ├─── *.raxml.bestTree  ──────────────────────────► Phase 2c (ASTRAL species tree)
    └─── *.raxml.support   ──────────────────────────► Phase 3  (D-statistic ML trees)
                                                       Phase 3  (D-statistic Bayesian trees) ◄─ Phase 2b
```

**Cluster:** GLBRC scarcity-ap-1 | **Conda env:** `phylo_env` | **User:** narhmadey

---

## Research Questions

1. **How has the genetic basis for mannose metabolism evolved across *Saccharomycotina*?**
   Maximum likelihood trees (RAxML-NG) and time-calibrated Bayesian trees (BEAST2) with MRCA anchors reveal when key evolutionary splits occurred across 267 species.

2. **Does the evolutionary history of these genes predict mannose growth ability?**
   Binary trait data (`1` = growth, `0` = no growth) mapped onto BEAST2 chronograms identifies eras where mannose metabolism was gained or lost.

3. **Does mannose growth show phylogenetic signal across gene trees?**
   The D-statistic (Fritz & Purvis 2010) tests whether the trait is phylogenetically structured or randomly distributed.

---

## Dataset Summary

```{r dataset-summary}
# Read the mannose traits CSV
# Adjust path relative to your working directory
traits_file <- "data/mannose_traits.csv"

if (file.exists(traits_file)) {
  trait_data <- read.table(traits_file,
                           header = TRUE, sep = "\t",
                           stringsAsFactors = FALSE,
                           quote = "", comment.char = "")

  cat(sprintf("Total species:        %d\n", nrow(trait_data)))
  cat(sprintf("Mannose growers (1):  %d\n", sum(trait_data$Mannose_Class == 1)))
  cat(sprintf("Non-growers (0):      %d\n", sum(trait_data$Mannose_Class == 0)))

  trait_data |>
    count(Mannose_Class) |>
    mutate(Phenotype = ifelse(Mannose_Class == 1, "Mannose grower", "Non-grower"),
           Percentage = round(100 * n / sum(n), 1)) |>
    select(Phenotype, Count = n, `%` = Percentage) |>
    kable(caption = "Mannose growth phenotype distribution across 267 yeast species") |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
} else {
  message("Trait file not found — place mannose_traits.csv in data/ to run this block.")
}
```

```{r dataset-plot, eval=file.exists("data/mannose_traits.csv")}
ggplot(trait_data, aes(x = factor(Mannose_Class,
                                  labels = c("Non-grower", "Mannose grower")),
                       fill = factor(Mannose_Class))) +
  geom_bar(width = 0.5, colour = "white") +
  scale_fill_manual(values = c("#d73027", "#4575b4"), guide = "none") +
  labs(title = "Mannose growth phenotype across 267 yeast species",
       x = NULL, y = "Number of species") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Environment Setup

The conda environment `phylo_env` must be created and activated before running any
pipeline steps on the cluster.

```{bash env-setup, eval=FALSE}
conda create -n phylo_env -c bioconda -c conda-forge \
    iqtree mafft trimal fasttree seqkit \
    raxml-ng beast2 beagle-lib \
    aster \
    r-base r-caper r-picante r-phytools r-ape

conda activate phylo_env

# Verify key tools
mafft --version
trimal -h | head -n 1
iqtree3 --version || iqtree --version
raxml-ng --version

# Package environment for Condor execute nodes
conda pack -n phylo_env -o phylo_env.tar.gz
```

---

## Preprocessing Pipeline (Local)

These steps run locally (or on the submit node) to produce the trimmed orthogroup
alignments (`OG*.trim.aln.fasta`) used by all downstream cluster analyses.

### Step A — Subset OrthoFinder FASTAs to 267 Species

Each full OG FASTA (1,154 species) is filtered to retain only sequences from the
267 focal species using `seqkit grep`.

> **Header note:** OG FASTA headers follow the pattern `>g009075.m1|aspergillus_nidulans.final`.
> The species list must contain the `*.final` IDs — the portion after the pipe.

```{bash subset-fastas, eval=FALSE}
# og_fastas_267.sh — run from project root
bash scripts/og_fastas_267.sh

# Verify outputs
ls -1 data/processed/03_ml_orthogroups/05_og_fastas_267/*.fasta | wc -l
head -n 20 data/processed/03_ml_orthogroups/logs/og_counts_267.tsv
```

```{r subset-qc, eval=FALSE}
# QC: Check sequence counts per OG after subsetting
counts <- read.table(
  "data/processed/03_ml_orthogroups/logs/og_counts_267.tsv",
  header = TRUE, sep = "\t"
)

counts |>
  arrange(Nseq_267) |>
  kable(caption = "Sequence counts per orthogroup after subsetting to 267 species") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Step B — Multiple Sequence Alignment (MAFFT)

Each orthogroup is aligned independently. OGs with fewer than 3 sequences are skipped.

| OG size | MAFFT mode | Rationale |
|---------|-----------|-----------|
| ≤ 200 sequences | `--localpair --maxiterate 1000` (L-INS-i) | High accuracy |
| > 200 sequences | `--auto` (FFT-NS-2) | Scalable; guide trees built twice |

```{bash mafft-run, eval=FALSE}
bash scripts/mafft_run.sh

# Verify
ls -1 data/processed/03_ml_orthogroups/06_alignments/*.aln.fasta | wc -l
tail -n 20 data/processed/03_ml_orthogroups/logs/mafft_run.log
```

### Step C — Alignment Trimming (TrimAl)

`trimal -automated1` adapts trimming thresholds per alignment — ideal for a batch
pipeline across OGs with widely varying length, divergence, and gap patterns. This
removes noisy columns introduced by the fast FFT-NS-2 aligner used on large OGs.

```{bash trimal-run, eval=FALSE}
bash scripts/trim_align.sh

# Verify
ls -1 data/processed/03_ml_orthogroups/07_trimmed_alignments/*.trim.aln.fasta | wc -l
tail -n 20 data/processed/03_ml_orthogroups/logs/trimal_run.log
```

### Step D — Local IQ-TREE Validation Test

A local IQ-TREE run validates the preprocessing pipeline on one orthogroup before
committing to the full cluster batch. Full model selection and tree inference for
all 9 OGs runs on the cluster (Phase 1–2).

```{bash iqtree-local, eval=FALSE}
bash scripts/iqtree_run.sh

ls -lah data/processed/03_ml_orthogroups/08_gene_trees/
tail -n 20 data/processed/03_ml_orthogroups/logs/iqtree_run.log
```

---

## Cluster Analysis Pipeline (GLBRC HT Condor)

```
Phase 1  —  Model Selection + IQ-TREE gene trees  (Condor, 9 parallel jobs, ~2 hrs)
Phase 2a —  ML Trees (RAxML-NG)                   (Condor, 9 parallel jobs, ~2 hrs each)
Phase 2b —  Bayesian Trees (BEAST2)               (Condor, 9 parallel jobs, ~24 hrs each)
Phase 2c —  Species Tree (ASTRAL)                 (submit node, after Phase 2a, ~minutes)
Phase 3  —  D-statistic                           (submit node, after Phase 2)
```

> **Phase 1 produces two outputs used downstream:**
> 1. `jobs_with_models.txt` — best-fit model per OG, passed directly to RAxML-NG (`--model`) in Phase 2a
> 2. `iqtree_logs/OG*.trim.aln_model.treefile` — IQ-TREE ML gene trees, used as BEAST2 starting trees in Phase 2b (via `generate_beast_xml.py`)

### Phase 1 — Model Selection (IQ-TREE ModelFinder, Condor parallel)

All 9 OGs are submitted simultaneously. Runtime: ~1.5 hrs (vs. ~13 hrs serial).

```{bash phase1-submit, eval=FALSE}
conda activate phylo_env
cd /home/glbrc.org/narhmadey/2.Y1000_267_phylogenetics/2_Plant_Path_563

# Generate alignment job list
ls data/OG*.trim.aln.fasta | xargs -n1 basename > scripts/modelfinder_jobs.txt
cat scripts/modelfinder_jobs.txt    # verify all 9 OGs listed

# Submit all 9 ModelFinder jobs simultaneously
condor_submit scripts/modelfinder_jobs.sub
condor_q                            # monitor — wait for all jobs to complete

# After completion: collect best models
bash scripts/collect_models.sh
cat scripts/jobs_with_models.txt    # verify all 9 models present
```

```{r phase1-results}
# Display completed model selection results
models <- data.frame(
  Orthogroup = c("OG0000007","OG0000008","OG0000009","OG0000020",
                 "OG0000029","OG0000032","OG0000044","OG0000067","OG0007179"),
  Best_Model = c("Q.YEAST+F+R10","LG+F+R10","LG+R10","Q.YEAST+R10",
                 "LG+F+R9","LG+F+R9","LG+F+R10","LG+F+R9","REMOVED"),
  Notes = c(
    "Yeast-specific matrix + empirical freqs + 10 FreeRate categories",
    "General LG + empirical freqs + 10 FreeRate categories",
    "General LG + 10 FreeRate categories",
    "Yeast-specific matrix + 10 FreeRate categories",
    "General LG + empirical freqs + 9 FreeRate categories",
    "General LG + empirical freqs + 9 FreeRate categories",
    "General LG + empirical freqs + 10 FreeRate categories",
    "General LG + empirical freqs + 9 FreeRate categories",
    "No outgroup (Lipomyces starkeyi absent); < 4 genes; excluded"
  )
)

models |>
  kable(col.names = c("Orthogroup", "Best Model (BIC)", "Model Components"),
        caption = "IQ-TREE ModelFinder results — best-fit substitution models (BIC criterion)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) |>
  row_spec(which(models$Orthogroup == "OG0007179"),
           color = "grey", italic = TRUE)
```

### Phase 2a — Maximum Likelihood Trees (RAxML-NG)

Each job: 10 CPUs, 8 GB RAM. Settings: 10 parsimony + 10 random starting trees,
1,000 bootstrap replicates, per-OG model from `jobs_with_models.txt`, seed `12345`.

```{bash phase2a-submit, eval=FALSE}
cat scripts/jobs_with_models.txt        # confirm all 8 models present

condor_submit scripts/raxml_jobs.sub    # submit all 8 ML jobs
condor_q                                # monitor
```

```{r phase2a-check, eval=file.exists("results")}
# Check which RAxML result files exist
raxml_files <- list.files("results", pattern = "\\.raxml\\.support$", full.names = FALSE)

if (length(raxml_files) > 0) {
  cat(sprintf("Found %d RAxML support tree(s):\n", length(raxml_files)))
  cat(paste0("  ", raxml_files, "\n"))
} else {
  cat("No RAxML support trees found yet in results/ — check condor_q\n")
}
```

### Phase 2c — Species Tree Inference (ASTRAL)

ASTRAL (run via the `aster` conda package) estimates a multi-species coalescent
species tree from the individual RAxML gene trees. It uses the `.raxml.bestTree`
files (unrooted, no bootstrap labels) rather than `.raxml.support`, because ASTRAL
uses branch lengths in coalescent units and support values on node labels would be
misread as taxon names.

This is the **coalescent-based counterpart** to the concatenation approach. While
RAxML and BEAST2 treat all genes as a single locus, ASTRAL explicitly models
incomplete lineage sorting (ILS) — gene tree discordance caused by deep ancestral
polymorphism — and is more appropriate when your gene trees show significant
topological conflict with each other.

```{bash astral-submit, eval=FALSE}
conda activate phylo_env

# Confirm ASTRAL is installed (from the aster package)
astral --help 2>&1 | head -n 3

# Run ASTRAL on all RAxML best trees
bash scripts/run_astral.sh

# Species tree written to: results/astral/astral_species_tree.tre
cat results/astral/astral_species_tree.tre
```

```{r astral-check, eval=file.exists("results/astral/astral_species_tree.tre")}
# Load and summarise the ASTRAL species tree
astral_tree <- read.tree("results/astral/astral_species_tree.tre")

cat(sprintf("ASTRAL species tree loaded\n"))
cat(sprintf("  Tips:       %d\n", length(astral_tree$tip.label)))
cat(sprintf("  Rooted:     %s\n", is.rooted(astral_tree)))
cat(sprintf("  Ultrametric:%s\n", is.ultrametric(astral_tree)))
```

```{r astral-plot, eval=file.exists("results/astral/astral_species_tree.tre")}
astral_tree <- read.tree("results/astral/astral_species_tree.tre")

# Basic ASTRAL species tree plot
plot(astral_tree,
     type       = "phylogram",
     cex        = 0.35,
     no.margin  = TRUE,
     main       = "ASTRAL multi-species coalescent tree")
```

```{r gene-tree-concordance, eval=file.exists("results") && file.exists("results/astral/astral_species_tree.tre")}
# Compare gene tree topologies to the ASTRAL species tree
raxml_trees <- lapply(
  list.files("results", pattern = "\\.raxml\\.bestTree$", full.names = TRUE),
  read.tree
)

if (length(raxml_trees) > 1) {
  # Count how many gene trees match the species tree topology on each node
  # (basic concordance check using Robinson-Foulds distances)
  rf_distances <- sapply(raxml_trees, function(gt) {
    tryCatch(RF.dist(astral_tree, gt, normalize = TRUE), error = function(e) NA)
  })

  og_names <- sub(".*/(OG[^.]+).*", "\\1",
                  list.files("results", pattern = "\\.raxml\\.bestTree$",
                             full.names = TRUE))

  rf_df <- data.frame(
    Orthogroup = og_names,
    RF_distance_normalised = round(rf_distances, 3),
    Concordance = ifelse(rf_distances < 0.2, "High",
                  ifelse(rf_distances < 0.5, "Moderate", "Low"))
  )

  rf_df |>
    kable(caption = "Normalised Robinson-Foulds distance: each gene tree vs. ASTRAL species tree
          (0 = identical topology, 1 = maximally discordant)") |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
    column_spec(3,
                color = ifelse(rf_df$Concordance == "High",   "#1a9850",
                        ifelse(rf_df$Concordance == "Moderate","#f46d43", "#d73027")))
}
```

### Phase 2b — Bayesian Inference (BEAST2)

Runs in parallel with Phase 2a. Settings: relaxed lognormal clock, Birth-Death
tree prior, 50 million MCMC steps, 2 independent runs per OG for convergence checking.

**MRCA Calibration Priors** (from Shen et al. 2018, Table S2):

```{r mrca-table}
mrca <- data.frame(
  Anchor = c("Deep", "Medium", "Young"),
  Taxa_Pair = c(
    "Babjeviella inositovora & Pachysolen tannophilus",
    "Zygosaccharomyces rouxii & Z. kombuchaensis",
    "Kazachstania siamensis & K. unispora"
  ),
  Mean_Ma = c(224.70, 36.90, 17.27),
  Lower_Ma = c(190.09, 22.98, 12.36),
  Upper_Ma = c(265.39, 51.96, 23.83)
)

mrca |>
  kable(col.names = c("Anchor", "Taxa Pair", "Mean (Ma)", "Lower (Ma)", "Upper (Ma)"),
        caption = "MRCA calibration priors for BEAST2 molecular clock (Shen et al. 2018)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```

```{bash phase2b-submit, eval=FALSE}
# Step 1: Generate XML configs (2 runs per OG)
python3 scripts/generate_beast_xml.py

# Step 2: Manually inject MRCA calibration priors into each beast_xml/*.xml
# (see README for XML code blocks)

# Step 3: Generate job list and submit
ls beast_xml/*.xml | xargs -n1 basename | sed 's/\.xml//' > scripts/beast_jobs.txt
condor_submit scripts/beast_jobs.sub
condor_q    # monitor — expect ~24 hrs per job

# Step 4: After jobs complete — check convergence in Tracer (ESS > 200)
# Step 5: Summarize posterior trees
for f in results/beast/*.trees; do
  og=$(basename $f .trees)
  treeannotator -burnin 10 $f results/beast/${og}.beast.tree
done
```

---

## Phase 3 — D-statistic Analysis

The D-statistic (Fritz & Purvis 2010) is calculated using `caper::phylo.d()` in R —
the same method used in the Y1000+ paper. It measures whether mannose growth
is phylogenetically structured or randomly distributed across gene trees.

### Load and Validate Trait Data

```{r load-traits}
traits_file <- "data/mannose_traits.csv"

if (!file.exists(traits_file)) {
  stop("Trait file not found: place mannose_traits.csv in data/ before running Phase 3.")
}

trait_data <- read.table(traits_file,
                         header = TRUE, sep = "\t",
                         stringsAsFactors = FALSE,
                         quote = "", comment.char = "")

# Validate required columns
required_cols <- c("Species", "Species_name_in_OGs", "Mannose_Class")
missing_cols  <- required_cols[!required_cols %in% colnames(trait_data)]
if (length(missing_cols) > 0) {
  stop(paste("Missing columns:", paste(missing_cols, collapse = ", ")))
}

# Validate binary trait
invalid <- unique(trait_data$Mannose_Class)[!unique(trait_data$Mannose_Class) %in% c(0, 1)]
if (length(invalid) > 0) {
  stop(paste("Non-binary values in Mannose_Class:", paste(invalid, collapse = ", ")))
}

cat(sprintf("Species loaded:       %d\n", nrow(trait_data)))
cat(sprintf("Mannose growers (1):  %d\n", sum(trait_data$Mannose_Class == 1)))
cat(sprintf("Non-growers (0):      %d\n", sum(trait_data$Mannose_Class == 0)))
```

### Load Gene Trees

```{r load-trees, eval=file.exists("results")}
raxml_trees <- list.files("results",
                          pattern    = "\\.raxml\\.support$",
                          full.names = TRUE)
beast_trees <- list.files(file.path("results", "beast"),
                          pattern    = "\\.beast\\.tree$",
                          full.names = TRUE)

all_trees <- c(
  setNames(raxml_trees,
           paste0(sub(".*/(OG[^.]+).*", "\\1", raxml_trees), "_ML")),
  setNames(beast_trees,
           paste0(sub(".*/(OG[^.]+).*", "\\1", beast_trees), "_Bayesian"))
)

cat(sprintf("Tree files found: %d\n", length(all_trees)))
for (nm in names(all_trees)) cat(sprintf("  %s\n", nm))
```

### Calculate D-statistic

```{r d-statistic, eval=file.exists("results")}
set.seed(42)

# Helper: interpret D value
interpret_d <- function(D, p_random, p_brownian) {
  if (is.na(D))             return("FAILED")
  if (p_random > 0.05)      return("No significant phylogenetic signal (p > 0.05)")
  if (D < 0)                return("Highly clustered — stronger than Brownian motion")
  if (D < 0.5)              return("Clustered — strong phylogenetic signal")
  if (D < 1.0)              return("Moderate phylogenetic signal")
  if (abs(D - 1.0) < 0.1)  return("Random — no phylogenetic signal")
  return("Overdispersed — convergent evolution")
}

results <- list()

for (tree_name in names(all_trees)) {
  tree_file <- all_trees[[tree_name]]
  og        <- sub("_(ML|Bayesian)$", "", tree_name)
  tree_type <- sub(".*_(ML|Bayesian)$", "\\1", tree_name)

  cat(sprintf("\n── %s (%s) ──────────────────────\n", og, tree_type))

  # Load tree
  tree <- tryCatch(read.tree(tree_file),
                   error = function(e) { cat("  ERROR reading tree\n"); NULL })
  if (is.null(tree)) next

  # Make binary and ultrametric
  if (!is.binary(tree))     tree <- multi2di(tree)
  if (!is.ultrametric(tree)) {
    tree <- tryCatch(chronos(tree, quiet = TRUE),
                     error = function(e) chronos(tree, model = "correlated", quiet = TRUE))
  }

  # Match trait names to tree tips
  csv_names   <- trait_data$Species_name_in_OGs
  n_matched   <- sum(csv_names %in% tree$tip.label)
  pct_matched <- round(100 * n_matched / length(csv_names), 1)

  cat(sprintf("  Tree tips:   %d | CSV species: %d | Matched: %d (%.1f%%)\n",
              length(tree$tip.label), length(csv_names), n_matched, pct_matched))

  if (n_matched < 10) {
    cat("  SKIPPED — fewer than 10 matched taxa\n")
    next
  }

  # Build comparative data
  named_trait  <- setNames(as.integer(trait_data$Mannose_Class),
                            trait_data$Species_name_in_OGs)
  matched_obj  <- tryCatch(match.phylo.data(tree, named_trait), error = function(e) NULL)
  if (is.null(matched_obj)) next

  comp_df <- data.frame(og_name = names(matched_obj$data),
                        mannose = as.integer(matched_obj$data),
                        row.names = names(matched_obj$data),
                        stringsAsFactors = FALSE)

  comp_data <- tryCatch(
    comparative.data(phy = matched_obj$phy, data = comp_df,
                     names.col = "og_name", vcv = TRUE, warn.dropped = FALSE),
    error = function(e) NULL
  )
  if (is.null(comp_data)) next

  # Run D-statistic
  d_result <- tryCatch(
    phylo.d(data = comp_data, binvar = mannose, permut = 1000),
    error = function(e) NULL
  )
  if (is.null(d_result)) next

  D_val      <- d_result$DEstimate
  p_random   <- d_result$Pval1
  p_brownian <- d_result$Pval0

  cat(sprintf("  D = %.4f | p(random) = %.4f | p(Brownian) = %.4f\n",
              D_val, p_random, p_brownian))
  cat(sprintf("  → %s\n", interpret_d(D_val, p_random, p_brownian)))

  results[[tree_name]] <- data.frame(
    Orthogroup    = og,
    Tree_Type     = tree_type,
    N_Tips        = length(tree$tip.label),
    N_Matched     = n_matched,
    Pct_Matched   = pct_matched,
    D             = round(D_val, 4),
    p_random      = round(p_random, 4),
    p_brownian    = round(p_brownian, 4),
    Interpretation = interpret_d(D_val, p_random, p_brownian),
    stringsAsFactors = FALSE
  )
}
```

### Results Summary

```{r results-table, eval=exists("results") && length(results) > 0}
results_df <- do.call(rbind, results)
rownames(results_df) <- NULL

# Write to file
write.table(results_df, "results/dstatistic_results.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)
cat("Results written to: results/dstatistic_results.tsv\n")

# Display
results_df |>
  select(Orthogroup, Tree_Type, N_Matched, Pct_Matched, D, p_random, p_brownian) |>
  kable(caption = "D-statistic results — phylogenetic signal for mannose growth trait",
        digits = 4) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) |>
  column_spec(5, bold = TRUE,
              color = ifelse(results_df$D < 0, "#d73027",
                      ifelse(results_df$D > 1, "#4575b4", "black")))
```

### D-statistic Visualisation

```{r d-plot, eval=exists("results_df") && nrow(results_df) > 0}
ggplot(results_df, aes(x = Orthogroup, y = D, colour = Tree_Type, shape = Tree_Type)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.6) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey70", linewidth = 0.6) +
  geom_point(size = 4, alpha = 0.85) +
  annotate("text", x = Inf, y = 0,  label = "D=0 (Brownian)",
           hjust = 1.05, vjust = -0.4, size = 3, colour = "grey40") +
  annotate("text", x = Inf, y = 1,  label = "D=1 (Random)",
           hjust = 1.05, vjust = -0.4, size = 3, colour = "grey70") +
  scale_colour_manual(values = c(ML = "#4575b4", Bayesian = "#d73027")) +
  labs(title = "D-statistic per orthogroup",
       subtitle = "Phylogenetic signal for mannose growth trait",
       x = "Orthogroup", y = "D-statistic",
       colour = "Tree type", shape = "Tree type") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"),
        legend.position = "top")
```

---

## D-statistic Interpretation Guide

```{r interp-table, echo=FALSE}
interp <- data.frame(
  D_Value = c("D < 0", "D = 0", "0 < D < 1", "D = 1", "D > 1"),
  Biological_Interpretation = c(
    "Highly clustered — stronger signal than Brownian motion",
    "Brownian motion — strong phylogenetic signal",
    "Moderate phylogenetic signal",
    "Random — no phylogenetic signal",
    "Overdispersed — convergent evolution"
  )
)

interp |>
  kable(col.names = c("D value", "Biological interpretation"),
        caption = "D-statistic interpretation (Fritz & Purvis 2010)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(1, background = "#fee8c8") |>
  row_spec(2, background = "#e0f3db") |>
  row_spec(5, background = "#deebf7")
```

**P-value guide:**

- `p(random) < 0.05` → trait is non-randomly distributed (phylogenetic signal present)
- `p(Brownian) < 0.05` → trait deviates from Brownian motion evolution

> **Note:** D-statistic on gene trees identifies which orthogroups are *associated*
> with mannose growth. D on the species tree (planned) asks whether mannose growth
> is broadly conserved across all 267 species.

---

## Planned Next Analyses

```{r next-steps, echo=FALSE}
next_steps <- data.frame(
  Step = 1:7,
  Analysis = c(
    "ASTRAL species-tree inference from ML gene trees",
    "Per-site conservation scoring",
    "Grower vs. non-grower enrichment tests",
    "Gene family presence/absence modelling",
    "Ancestral state reconstruction",
    "Mannose metabolism pathway mapping",
    "Genome alignment (pygenomeviz/progressiveMauve vs. Cactus)"
  ),
  Status = c("Integrated (Phase 2c)","Pending","Pending","Pending","Pending","Pending","In progress")
)

next_steps |>
  kable(caption = "Planned downstream analyses") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  column_spec(3, color = ifelse(next_steps$Status == "In progress", "#e08214", "grey50"))
```

---

## Reproducibility

```{r session-info}
sessionInfo()
```

```{r tool-versions, eval=FALSE}
# Run on the cluster to capture tool versions
system("mafft --version")
system("trimal -h 2>&1 | head -n 2")
system("iqtree3 --version 2>&1 | head -n 2")
system("raxml-ng --version 2>&1 | head -n 2")
```

---

*Project: PLANTPATH 563 | GLBRC scarcity-ap-1 | Conda env: `phylo_env` | Last updated: March 2026*
